---
title: "My NYPD Shooting Incident Data Report"
author: "Y. Habib"
date: "2023-12-11"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(leaflet)
library(leaflet.extras)
library(dplyr)
library(ggplot2)
library(tidyr)
library(lubridate)
library(forecast)
```


```{r Importing_and_Cleaning_Data}

url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
NYPDData <- read.csv(url)

summary(NYPDData)

NYPDShootingData <- NYPDData

#transforming date from <chr> to <date>
NYPDShootingData$OCCUR_DATE <- as.Date(NYPDShootingData$OCCUR_DATE, format = "%m/%d/%Y")

print(head(NYPDShootingData))


#Checking to see if there are columns with empty cells
empty_cells_per_column <- colSums(NYPDShootingData == "")
print(empty_cells_per_column)

```
A large number of location description cells are empty which raises the following question:
Is there a discernible pattern in the allocation of resources for shooting investigations
resulting in the lack of documentation?

Missing data can potentially indicate bias...

General possible sources of bias:

- Bias by law enforcement such as racial profiling

- Bias from missing data 

- Bias from different practices in different Boroughs (possibly economic disparities)

- Bias from what economic indicators I chose to use...maybe if I used different indicators
  I would make different observations

Since there is a large number of shootings with missing cells, they will be kept in the
data set BUT the missing data will be further investigated based off the question above
at the end of the document.






Question: Are there certain locations that have a higher frequency of incidents?


```{r Further_Cleaning_Data}

#Removing all columns that serve no purpose in answering the question above
NYPDShootingData <- NYPDShootingData %>% select(-INCIDENT_KEY, -STATISTICAL_MURDER_FLAG, 
                                                -PERP_AGE_GROUP, -PERP_SEX, -PERP_RACE, 
                                                -VIC_AGE_GROUP, -VIC_SEX, -VIC_RACE)

print(head(NYPDShootingData))

```


```{r Data_Visualization}

#Using bar graphs for better distiction of the different boroughs:
ggplot(NYPDShootingData, aes(x = BORO)) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(title = "Distribution of Shooting Incidents across Neighborhoods",
       x = "Neighborhood",
       y = "Number of Incidents") +
  theme_minimal()

```

...

Questions Raised:

1.) Has there been a change in the distribution of shootings across various boroughs 
    over the years?
    
2.) How has number of shootings changed over the years

...

Distribution of Shootings Across Boroughs:


```{r Dist_of_Shootings_Across_Boroughs}

NYPDShootingData$Year <- year(NYPDShootingData$OCCUR_DATE)

# Group by year and borough
shooting_by_year <- NYPDShootingData %>%
  group_by(Year, BORO) %>%
  summarise(Incidents = n())

# Create a line chart
ggplot(shooting_by_year, aes(x = Year, y = Incidents, color = BORO)) +
  geom_line() +
  labs(title = "Distribution of Shootings Across Boroughs Over Years",
       x = "Year",
       y = "Number of Incidents",
       color = "Neighborhood") +
  theme_minimal()

```

Observation 1: The poorer boroughs of NY consistently have a higher number of incidents
than the richer boroughs

~This can potentially simply be because of population density~

Further investigation needed to confirm...

Observation 2: Number of incidents in each borough have similar trends (generally
decreasing until 2020 when a large spike in incidents is seen)

...

Number of Overall Shootings per Year:

``` {r Number_of_Shootings_per_Year}

shootings_per_year <- NYPDShootingData %>%
  group_by(Year) %>%
  summarise(Incidents = n())

print(shootings_per_year)

ggplot(shootings_per_year, aes(x = Year, y = Incidents)) +
  geom_line() +
  labs(title = "Number of Shootings Per Year",
       x = "Year",
       y = "Number of Incidents") +
  theme_minimal()

```

Observation: Total number of incidents follows the same trend observed in individual
boroughs (generally decreasing, large spike in 2020)

Question Raised:
- Does the number of shootings exhibit dependence on economic conditions?

No single indicator can fully capture the complexity of the economy, after some research,
two indicators will be used: inflation rate, and unemployment rate.

...

Inflation Rate:

```{r Shootings_and_Economy_Inflation_Rates}

#https://fred.stlouisfed.org/series/FEDFUNDS
USInflationRates <- read.csv("~/Desktop/united-states-inflation-rate-cpi.csv", skip = 16, header = TRUE)

#Removing empty "X" column
USInflationRates <- USInflationRates %>% select(-X)

#Changing date column to date format
USInflationRates$date <- as.Date(USInflationRates$date, format = "%Y-%m-%d")

#Removing all dates that are before 2006-01-01 since that is the beginning of the shooting data
USInflationRates <- USInflationRates %>%
  filter(date >= as.Date("2006-01-01"))

#Creating a new column of the year
USInflationRates$Year <- year(USInflationRates$date)

print(head(USInflationRates))

#Create a graph of Inflation rate over the years
ggplot(USInflationRates, aes(x = Year, y = Inflation.Rate....)) +
  geom_line() +
  labs(title = "Inflation Rate per Year",
       x = "Year",
       y = "Inflation Rate") +
  theme_minimal()

```

Inflation rate spikes at 2021

...

Unemployment Rate:

```{r Shootings_and_Economy_Unemployment_Rates}

#https://fred.stlouisfed.org/series/UNRATE
USUnemploymentRates <- read.csv("~/Desktop/UNRATE.csv")

#Changing date column to date format
USUnemploymentRates$DATE <- as.Date(USUnemploymentRates$DATE, format = "%Y-%m-%d")

#Removing all dates before 2006-01-01 since that is the beginning of the shooting data
USUnemploymentRates <- USUnemploymentRates %>%
  filter(DATE >= as.Date("2006-01-01"))

print(head(USUnemploymentRates))

ggplot(USUnemploymentRates, aes(x = DATE, y = UNRATE)) +
  geom_line() +
  labs(title = "Unemployment Rate per Year",
       x = "Year",
       y = "Unemployment Rate") +
  theme_minimal()

```

General downwards trend but with two spikes seen around 2008 and 2020


Conclusion: There seems to be a trend between the wealth in the region and the number
of shooting incidents, specifically the wealthier the borough, the less shooting incidents
occur there. There also seems to be a trend in how the US economy is doing and the 
shooting incident rate which leads me to conclude that the better off an area is 
financially, the less shootings there will be. 
Further research is needed to make a more solid and evidence based conclusion but with 
the limited knowledge I have now, this was the best I can do.

...

MODELING:

Predicting what shooting incidents will look like in the future...


```{r Modeling}

monthly_counts <- NYPDShootingData %>%
  filter(!is.na(OCCUR_DATE)) %>%
  group_by(year_month = format(OCCUR_DATE, "%Y-%m")) %>%
  summarize(Number_of_Incidents = n())

shooting_time_series <- ts(monthly_counts$Number_of_Incidents, frequency = 12, start = c(as.numeric(substr(monthly_counts$year_month[1], 1, 4)), as.numeric(substr(monthly_counts$year_month[1], 6, 7))))

shooting_forecast <- forecast(auto.arima(shooting_time_series), h = 12)

autoplot(shooting_forecast) + ggtitle("Monthly Shooting Incidents Forecast")

```

This prediction model predicts that incidents will continue to spike and dip. This 
analysis does not take into consideration economic predictions. Further analysis is 
required to better the model but with the limited knowledge we have now, this was the 
best I can do.

...

ANALYZING MISSING DATA:

```{r Missing_Data}

#New table with only rows that contain missing cells
NYPDDataWithMissing <- NYPDData[rowSums(NYPDData == "") > 0, ]

#New table that counts the number of shootings in each Borough
Missing_Shooting_Summary <- NYPDDataWithMissing %>%
  group_by(BORO) %>%
  summarise(Shootings = n())

Missing_Shooting_Summary <- Missing_Shooting_Summary %>%
  rename(Shootings_with_Missing = Shootings)

print(Missing_Shooting_Summary)


#Looking at the proportion of shootings in each borough that is incomplete:
Shooting_Summary <- NYPDData %>%
  group_by(BORO) %>%
  summarise(Shootings = n())

print(Shooting_Summary)

Merged_Shooting_Summary <- merge(Missing_Shooting_Summary, Shooting_Summary, by = "BORO", all.x = FALSE, all.y = FALSE)

Merged_Shooting_Summary$Prop_of_Missing <- Merged_Shooting_Summary$Shootings_with_Missing / Merged_Shooting_Summary$Shootings

```


Based off this analysis, Manhattan, which is a wealthier borough than all the rest, 
has a lower proportion of shootings with missing cells. This can potentially be a 
source of bias such as racial profiling or such as racial profiling or bias from 
different practices in different Boroughs possibly coming from economic disparities.
Further research and analysis is necessary...

Next steps possibly include running an analysis to see if the difference is statistically
significant.


